Deep Reinforcement Learning for Conservation Decisions

### 2021-04-20 manuscript/appendix.Rmd
:pen:

At the heart of each algorithm is a neural network. The type, size and depth of network, choice of activation functions and so forth, are considered "hyper-parameters" of algorithm: 
we must first make a selection for each of these hyper-parameters to define our (untrained) agent.  
Other hyper-parameters determine other aspects of the training regime: such as how the agent chooses to explore possible actions, the learning rate, or how future rewards are discounted (gamma).
Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
for instance, the A2C algorithm originally described in @A2C. 
Here, we train an agent using the A2C algorithm using default hyperparameter settings.

@@

Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyperparameter selection. This is discussed further below.

We can load a trained agent and use `simulate` and `policyfn` methods provided by the `gym_fishing` environment to simulate the process for a fixed number of replicates.

@@

@@

## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( 0.75 )

for(i in 1:100){

@@

@@

@@

We train a second agent based on the TD3 algorithm.  
This time, we will declare our choices for the hyper-parameters explicitly.
Note that some of these hyperparameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
Good choices for hyperparameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance. 
For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].

@@

Once again, after our agent is defined we are ready to train it.

@@

@@

@@

@@

```{r}
write_csv(sims_df, "figs/sims_df.csv")
write_csv(policy_df, "figs/policy_df.csv")
write_csv(reward_df, "figs/reward_df.csv")
```

@@

@@

>>>>

At the heart of each algorithm is a neural network. The type, size and depth of network, choice of activation functions and so forth, are considered "hyper-parameters" of algorithm: 
we must first make a selection for each of these hyper-parameters to define our (untrained) agent.  
Other hyper-parameters determine other aspects of the training regime: such as how the agent chooses to explore possible actions, the learning rate, or how future rewards are discounted (gamma).
Each algorithm may have different hyper-parameters. `stable-baselines3` provides default values for all hyperparameters based on the original papers that introduced the corresponding algorithms,
for instance, the A2C algorithm originally described in @A2C. 
Here, we train an agent using the A2C algorithm using default hyperparameter settings, utilizing a multi-layer perceptron neural network composed of two 64-neuron layers.

@@

Note that while default hyperparameters provide a useful starting place (particularly when the environment has been suitably normalized, a best-practice we discuss above), better performance can almost always be achieved by _tuning_ the hyperparameter selection. This is discussed further below. 
Having saved our trained agent to disk (`cache/a2c.zip`), we can then re-load this agent for evaluation or to continue training.  Note that a copy of the trained agents are included in the corresponding GitHub repository.

@@

We can supply an observation to our trained agent and it will return it's proposed action using the `predict` method. 
This is all we need to evaluate or employ the agent on a decision-making task.
Recall that state space and action space in the fishing gym have been re-scaled to a (-1, 1) interval.
Note that this is equivalent to a choice of appropriate units -- we can re-scale the interval without loss of generality.
This is often a critical step in the design of an RL environment to facilitate successful training.
Following the `gym` standard, the core methods such as `predict` and `step` operate on the re-scaled units,
so it is necessary to first transform the original units into this re-scaled state space.
For example, if we wish to start a simulation with a stock size of 0.75, we can use the helper
method `get_state()`, to determine the corresponding value in the re-scaled state space.  
Unlike `predict` and `step`, `get_state()` is not a standard method of all gym environments -- typically
a user must first inspect the state space of an environment and choose themselves how to re-scale their
problem into that state space.

@@

## represent the initial state size in the 'rescaled' state space.
state <- env$get_state( 0.75 )
state
```

With an initial state in hand, we are ready to simulate management using our agent.
The iteration is simple: we use the agent to predict what action we should take given the current state,
and then we take said action and examine the result to determine the future state. 
Because these methods return additional information as well, a little extra sub-setting is required in R:

```{r}
for(i in 1:10){

@@

For convenience, `gym_fishing` defines the helper routine `simulate` to perform the above iteration `reps` number of times. The `simulate` method returns the state, action, and reward resulting from each time step of each replicate:

@@

Because the agent typically attempts to retain the fish population near a specific value, simulations with well-trained agents will usually not explore the full range of possible states.
To get a better idea of how the agent behaves across the full state space, it common to consider a policy function: a plot of the action taken at each possible state.
Some agents are non-deterministic, so we may want to use replicate draws at each state to get a better picture of the agent's behavior.
Note that the `policyfn` method is another custom method and not part of the `gym` standard, and the use of a policy function will not make sense for agents that consider the history of many previous states in selecting their action, such as agents which utilize recurrent neural network architectures like LSTM [@LSTM].

@@

We train a second agent based on the TD3 algorithm. 
This illustrates some of the salient differences between the different algorithms used in reinfocement learning.
Bear in mind that the best algorithm for any given environment will depend on the details of the environment, as well
as the length of training and the selection of hyper-parameters, including the agent's underlying neural networks which provide the 'brains' of the agent's decision-making ability.
This time, we will declare our choices for the hyper-parameters explicitly.
Note that some of these hyper-parameters, such as the `learning_rate` and discount rate, `gamma`, are also hyper-parameters of the A2C algorithm, while others, like the `action_noise` describing how new actions are proposed, are not used in A2C.
Good choices for hyper-parameters can be found through a process known as "hyper-parameter tuning", which simply uses standard function optimization approaches to vary the choice of hyper-parameters used in training to determine which maximizes the agent's performance. 
For simplicity, we show only training with specified hyper-parameters here, example scripts for tuning the algorithm to discover these values can be found in <https://github.com/boettiger-lab/rl-toolkit>, and compare to the tuning utilities found in `stable-baselines3-zoo` [@zoo].

@@

Once again, after our agent is defined we are ready to train it:

@@

With our trained agent, we can simulate replicate scenarios or scan across states to determine the implicit policy function:

@@

We compare the performance of these two RL agents trained using the algorithms A2C and TD3 respectively to the known optimal policy for this particular environment.  
Recall that under the assumptions of the simple model used in the `fishing-v1` environment, we can determine the optimal harvest policy analytically if the model and parameters are known precisely [@Reed1979].
The optimal strategy is a policy of 'constant escapement', designed to keep the remaining stock size (the population that 'escapes' fishing harvest) at the biomass corresponding to a maximum growth rate, i.e. at $B = K/2$ in this model.
`gym_fishing` defines a collection of non-RL agents in the `models` submodule, including the a human agent that merely asks to enter their desired quota manually. 
The `escapement` model implements the provably optimal constant escapement rule.
A third model, `msy`, implements a policy based on "Maximum Sustainable Yield" policy [@Schaefer1954], which is actually more commonly used as a basis for management than constant escapement, despite only being optimal at the steady state under deterministic dynamics.

@@

Having performed simulations under each method, we gather the results under each management strategy together into a single data frame for simulations (`sims_df`), the policy function (`policy_df`) and the cumulative reward, used to make the plots in the corresponding panels.  
We preserve a copy of each data frames in local `.csv` text files for maximum cross-platform compatibility.

@@

write_csv(sims_df, "figs/sims_df.csv")
write_csv(policy_df, "figs/policy_df.csv")
write_csv(reward_df, "figs/reward_df.csv")
```

Standard R methods are used to plot the results summarized across the replicates:

@@

Our second example utilizes our `gym_conservation` to provide the necessary environment to simulate an ecosystem approaching a tipping point.
For simplicity we will train only a TD3-based agent this time.

@@

In general, the optimal solution depends on the ecological dynamics, the benefit of the ecosystem services and the costs associated with a management response.
Because the tipping point problem is non-autonomous, we cannot solve for the optimal policy even given the model and objective (utility) function using Markov Decision Process methods.
However, a simple heuristic solution provides a reasonable starting point for comparison.

As discussed in the main text, we have no existing optimal solution to the tipping point problem, and so rely on a common heuristic strategy instead:
select a fixed level of conservation investment that is sufficient to counter-balance any further side towards the tipping point, preserving it in it's current state.

### 2021-04-20 manuscript/manuscript.Rmd
:pen:

When the environment is fully observable, i.e. when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP). An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$. As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence. From this definition, we can provide an agent's trajectory distribution for a given policy as,
$$
  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
$$

@@

Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state. Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory. There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$. The state-value function is defined as the expected return from a given state,
$$
  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

@@

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach is in principle agnostic to the details of the simulation.

```{r}
sims_df <- read_csv( "figs/sims_df.csv")
policy_df <- read_csv("figs/policy_df.csv")
reward_df <- read_csv("figs/reward_df.csv")

@@

@@

## Example of solving a problem that is difficult/impossible for conventional methods

two use cases to consider: 
  - model is known but too complex to solve by (approximate) dynamic programming etc, or non-markovian
  - model is unknown

model uncertainty/transfer learning? nonstationary dynamics?

>>>>

When the environment is fully observable, i.e. when the state that the agent observes fully characterizes the system, the reinforcement learning problem can be formalized as a Markov Decision Process (MDP).
An MDP consists of the following: a set of states, $\mathcal{S}$, called the state space; a set of actions, $\mathcal{A}$, called the action space; a state transition operator, $T(s_{t+1}|s_t, a_t)$, which describes the dynamics of the system; a reward function, $r(s_t, a_t)$; an initial state distribution, $d_0(s_0)$; and a discount factor, $\gamma \in (0,1]$. 
As the agent interacts with the environment, the agent will create a trajectory in state-action space given as $\tau = (s_0, a_0, \dots, s_H, a_H)$ where H denotes the horizon or length of the state-action sequence.
From this definition, we can provide an agent's trajectory distribution for a given policy as,

$$
  p_\pi(\tau) = d_0(s_0) \prod_{t=0}^H \pi(a_t \mid s_t) T(s_{t+1} \mid s_t, a_t)
$$

@@

Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state. Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high reward trajectory.
There are 3 value functions that are commonly used across RL algorithms: the state value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$. The state-value function is defined as the expected return from a given state,

$$
  V^\pi(s_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg]
$$

@@

For illustrative purposes, we consider the simplest version of the dynamic optimal harvest problem as outlined by @Clark1974 (for the deterministic case) and @Reed1979 (under stochastic recruitment).
In short, the manager seeks to optimize the net present value (discounted cumulative catch) of a fishery, observing the stock size each year and setting an appropriate harvest quota in response.
In the classical approach, the best model of the fish population dynamics must first be estimated from data, potentially with posterior distributions over parameter estimates reflecting any uncertainty.
From this model, the optimal harvest policy -- that is, the a function which returns the optimal quota for each possible observed stock size -- can be determined either by analytic [@Reed1979] or numerical [@Marescot2013] methods, depending on the complexity of the model.
In a RL approach, the model is instead replaced by a computer simulation, usually referred to as an "environment."
The simplest simulation is merely an implementation of the estimated model itself.
However, unlike an analytic or SDP solution that is derived directly from the estimated model, the RL approach makes no assumption about the precise functional form or parameter values underlying the dynamics -- it is in principle agnostic to the details of the simulation.

To better illustrate the potential differences between different RL algorithms used to train the agents, we will examine two RL solutions, one based on A2C [@A2C] and the other based on @TD3.
A step-by-step walk-through for training both agents on this environment is provided in the Appendix.
We compare the resulting management, policy, and reward under either RL agent to that achieved by the optimal management solution [Fig 1].
Despite having no knowledge of the underlying model and clear differences in the resulting strategies, both RL agents learn enough to achieve nearly optimal performance.

```{r message=FALSE, include=FALSE}
sims_df <- read_csv( "figs/sims_df.csv")
policy_df <- read_csv("figs/policy_df.csv")
reward_df <- read_csv("figs/reward_df.csv")

@@

Our second example focuses on a case for which we do not already have an existing, provably optimal policy to compare against.
We consider the generic problem of an ecosystem facing slowly deteriorating environmental conditions which move the dynamics closer towards a tipping point [Fig 2].
This model of a critical transition has been posited widely in ecological systems, from the simple consumer-resource model of @May1977 on which our dynamics are based, to microbial dynamics [@Dai2013] lake ecosystem communities [@Carpenter2008] to planetary ecosystems [@Barnkosky2014].
On top of these ecological dynamics we introduce an explicit ecosystem service model quantifying the value of more desirable 'high' state relative to the 'low' state.
For simplicity, we assume a proportional benefit $b$ associated with the ecosystem state $x(t)$.
Thus when the ecosystem is near the 'high' equilibrium $\hat x_H$, the corresponding ecosystem benefit $b \hat x_H$ is higher than at the low equilibrium, $b x_L$, consistent with the intuitive description of ecosystem tipping points [@Barknosky2014]. 
We also enumerate the possible actions which a manager may take in response to environmental degradation. 
In the absence of any management response, we assume the environment deteriorates at a fixed rate $\alpha$, which can be thought of as the incremental increase in global mean temperature or similar anthropogenic forcing term.
Management can respond to slow or even reverse this trend by choosing an opposing action $a_t$.
We assume that large actions are proportionally more costly than small actions, consistent with the expectation of diminishing returns: taking the cost associated with an action $a_t$ as equal to $c a_t^2$.
Many alterations of these basic assumptions are also possible: our `gym_conservation` implements a range of different scenarios with user-configurable settings.
In each case, the manager observes the current state of the system each year and must then select the policy response that year.

Because this problem involves a parameter whose value changes over time (the slowly deteriorating environment), the resulting ecosystem dynamics are not autonomous.
This precludes our ability to solve for the optimal management policy using classical theory such as for Markov Decision Processes (MDP, @Marescot2013), typically used to solve sequential decision-making problems.
However, it is often argued that simple rules can achieve nearly optimal management of ecological conservation objectives in many cases [@Possingham; @Possingham; @morePossingham].
A common conservation strategy employs a fixed response level rather than a dynamic policy which is toggled up or down each year: for example, declaring certain regions as protected areas in perpetuity.
An intuitive strategy faced with an ecosystem tipping point would be 'perfect conservation', in which the management response is perfectly calibrated to counter-balance any further decline.
While the precise rate of such decline may not be known in practice (and will not be known to RL algorithms before-hand either), it is easy to implement such a policy in simulation for comparative purposes. 
We compare this rule-of-thumb to the optimal policy found by training an agent using the TD3 algorithm:

@@

The TD3-trained neural network agent proves far more successful in preventing chance transitions across the tipping point, consistently achieving a higher cumulative ecosystem service value across replicates than the steady-state strategy.

Examining the replicate management trajectories and corresponding rewards reveal that the RL agent incurs significantly higher costs in the initial phases of the simulation, dipping the cumulative reward into negative territory on average and well below the reward realized by the simple steady-state strategy.
This initial investment then begins to pay off -- by about 200th time step the RL agent has surpassed the performance of the steady state strategy.  
The policy plot provides more intuition for the RL agent's strategy: at very high state values, the RL agent opts for no conservation action -- so far from the tipping point, no response is required.
Near the tipping point, the RL agent steeply ramps up the conservation effort, and retains this effort even as the system falls below the critical threshold, where a sufficiently aggressive response can tip the system back into recovery.
For a system at or very close to the zero-state, the RL agent gives up, opting for no action.

A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy (see appendix, fig S1, S2) further highlights the difference between these two approaches.
Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards. 
In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery (Fig S2).
Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.

Such examination makes it clear how the RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
The various kinks in the RL policy function also suggest that this solution is nevertheless surely not the optimal solution -- such features are typical of RL-based policies, where such spandral features from early experimentation may become baked into the long-term strategy if adjusting them has little improvement on the final reward.
Further tuning of hyper-parameters, increased training, alterations to the training algorithm or an entirely different training algorithm would likely be able to further improve upon this performance.

In particular, the use of a recurrent neural network architecture instead of the multi-layer perceptron used here would open the door for the policy to reflect the non-autonomous dynamics more directly.
A recurrent neural network, such as long-short-term-memory (LSTM), could learn a strategy which reflects not only the current system state, but also latent information such as the time elapsed.
